Response to reviewers

### 2025-01-23 Manuscript/ecography_reponse_letter_2.tex
Update ecography_reponse_letter_2.tex

\textcolor{blue}{Recommendation by the Subject Editor (Dr. Barry Brook):}

@@

% - please focus on resolving these methodological and conceptual concerns to improve transparency and strengthen the presentation of your core findings.
% - You will also need to address the referees’ points regarding:
%   - the consistency of dimension usage,
%   - the rationale for simulating “multidimensional” traits,
%   - and the potential biases introduced by your second scaling step.

@@

\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g. \cite{novack2016general1,novack2016general2}), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines @@@.

@@

\textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more likely compared to common ones).} Lines @@@

@@

\textit{here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D} Lines @@@.

@@

\textit{Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines @@@

@@

\textit{For example, we can follow the suggestions from \cite{mammola2021concepts} highlighting which kind of methodological pipeline best supports which type of metric of interest (Table 3).} Lines @@@

@@

\textit{First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} \citep{rcore} using \texttt{treats} \citep{guillerme2024treats}.
We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines @@@

However, we kept the broader definition of trait.
\citealt{violle2007let} provided an ``unambiguous definition of plant trait, with a particular emphasis on functional trait'', which as per the authors aims was proposed to clarify and and conceptualise the definition on functional traits (``a surrogate of organismal performance'') specifically in plant ecology.
Here we build on \citealt{mcgill2006rebuilding}'s more generalised approach, and, more interestingly on the work of \citealt{dawson2021traits} of how ecologists _actually_ operationalise the concept of trait based on researchers understanding vs. proposed definitions.
We propose through the paragraph on lines @@@ to expand that to ``any measurable aspect of an organism'' which we hope allows expansion beyond one dimensional traits and, more importantly, beyond ecology.
I.e. we believe that this definition of trait more accurately also includes how traits are used in microbiology (e.g. bacterial colony shape), macroevolution (e.g. 3D geometric morphometrics) or in palaeontology (e.g. ``position of a specific foramen in a skull relative to another feature'').

% TODO: also define this in the new in-text glossary table.

@@

\textit{This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.} lines @@@

We now also mention it in the discussion:

\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature \citep{bellman1957dynamic}.} Lines @@@

@@

\textit{Note that our definition of a trait is a generalisation of \cite{violle2007let,mcgill2006rebuilding,dawson2021traits}'s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).} Lines @@@

>>>>

% \textcolor{blue}{Recommendation by the Subject Editor (Dr. Barry Brook):}

@@

% - please focus on resolving these methodological and conceptual concerns to improve transparency and strengthen the presentation of your core findings.

% - You will also need to address the referees’ points regarding:
%   - the consistency of dimension usage,
%   - the rationale for simulating “multidimensional” traits,
%   - and the potential biases introduced by your second scaling step.

@@

\textit{Although some previous work has been done in understanding patterns in the context of mechanisms and process (e.g. Novack-Gottshall (2016a,b))), we argue that often the pattern description follows some previously used framework without a proper evaluation of its adequacy to answer the how and why.} Lines 89-92. %@@@

@@

\textit{In other words, pairs of observations that share similar trait combinations are more common in the trait space than observations with dissimilar trait combinations (i.e. observations adjacent to each other in the trait space are more likely compared to common ones).} Lines 175-178. %@@@

@@

\textit{here we arbitrarily used $p=3$ which was a relatively low scaling value that still resulted in visible changes observable in 2D} Lines 202-203.% @@@.

@@

\textit{Our results show that the loss of  functional diversity for several metrics was not more than expected by chance.
This could be due to the taxonomic distribution of the extinct birds: this example data has approximatively $58$\% passerines while the majority of anthropogenic extinctions have been affecting non-passerines (approximatively $75$\%).
These species tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).} Lines 431-436. %@@@

@@

\textit{For example, we can follow the suggestions from Mammola et al. (2021) highlighting which kind of methodological pipeline best supports which type of metric of interest (Table 3).} Lines 545-547. %@@@

@@

\textit{First, we simulated multiple independent random time dependent trait (Brownian Motion) under a model where lineages only speciate (no extinction; i.e. a pure birth speciation model) until reaching 200 observations in \texttt{R} (R Core Team, 2024) using \texttt{treats} (Guillerme, 2024).
We simulated either 2, 4 or 8 independent (uncorrelated) Brownian Motion traits (equivalent to 2, 4 or 8 dimensional traits).} Lines 124-124. %@@@

However, we kept the broader definition of trait.
Violle et al. (2007) provided an ``unambiguous definition of plant trait, with a particular emphasis on functional trait'', which as per the authors aims was proposed to clarify and and conceptualise the definition on functional traits (``a surrogate of organismal performance'') specifically in plant ecology.
Here we build on McGill et al. (2006)'s more generalised approach, and, more interestingly on the work of Dawson et al. (2021) of how ecologists \textit{actually} operationalise the concept of trait based on researchers understanding vs. proposed definitions.
We propose through the paragraph on lines 124-139 to expand that to ``any measurable aspect of an organism'' which we hope allows expansion beyond one dimensional traits and, more importantly, beyond ecology.
I.e. we believe that this definition of trait more accurately also includes how traits are used in microbiology (e.g. bacterial colony shape), macroevolution (e.g. 3D geometric morphometrics) or in palaeontology (e.g. ``position of a specific foramen in a skull relative to another feature'').

@@

\textit{This curse changes the properties of space occupancy in a non-linear way depending on each observation's distribution and the number of dimensions.} lines 251-253.
We now also mention it in the discussion:

\textit{Note also that, as mentioned above, different dimensionalities can have different effects on the metrics, depending on both the traitspace and the metric nature (Bellman et al., 1957).} Lines388-390. %@@@

@@

\textit{Note that our definition of a trait is a generalisation ofViolle et al. (2007); McGill et al.
(2006); Dawson et al. (2021)’s suggestions which is expanded to include traits beyond the ones used for functional ecology (macroevolution, palaeontology, microbiology, etc.).} At the end of Table 1.
